---
title: "Technical Analysis Update: PAH Exposure Modeling Sensitivity and Primary Model"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## 1. Background and goals

This report summarizes the model-development workflow used to evaluate condition effects on total PAH exposure under sparse data and substantial non-detects.

The primary methodological decisions addressed were:

1. How to handle non-detects.
2. How many PAHs to include in the response definition.
3. How to handle repeated measures.
4. What model form to use for adjusted between-condition contrasts.

The sequence below follows the exploratory-to-primary progression:

- **Script 10**: non-imputed baseline boundary analysis.
- **Script 09**: sensitivity to imputation beta.
- **Script 11**: sensitivity to number of PAHs under consistent MLE imputation.
- **Script 12**: selected primary analysis.

## 2. Setup and artifact loading

```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)

# Script outputs
out_09 <- "09_output"
out_10 <- "10_output"
out_11 <- "11_output"
out_12 <- "12_output"

# ---- Script 10 artifacts ----
step10_2 <- readRDS(file.path(out_10, "step2_artifacts.rds"))
step10_3 <- readRDS(file.path(out_10, "step3_artifacts.rds"))
qc10 <- readRDS(file.path(out_10, "step2_model_qc.rds"))

cond10 <- step10_2$condition_contrast_results
time10 <- step10_2$timing_results

# ---- Script 09 artifacts ----
step09_2 <- readRDS(file.path(out_09, "step2_results.rds"))
cond09 <- readRDS(file.path(out_09, "condition_contrast_results.rds"))
time09 <- readRDS(file.path(out_09, "timing_results.rds"))

# ---- Script 11 artifacts ----
step11_6 <- readRDS(file.path(out_11, "step6_artifacts.rds"))
step11_7 <- readRDS(file.path(out_11, "step7_artifacts.rds"))
step11_8 <- readRDS(file.path(out_11, "step8_artifacts.rds"))

cond11 <- step11_6$condition_contrast_results
time11 <- step11_6$timing_results

# ---- Script 12 artifacts ----
step12_1 <- readRDS(file.path(out_12, "step1_artifacts.rds"))
step12_2 <- readRDS(file.path(out_12, "step2_artifacts.rds"))
step12_3 <- readRDS(file.path(out_12, "step3_artifacts.rds"))
step12_4 <- readRDS(file.path(out_12, "step4_artifacts.rds"))
step12_5 <- readRDS(file.path(out_12, "step5_artifacts.rds"))
step12_6 <- readRDS(file.path(out_12, "step6_artifacts.rds"))

fit12 <- step12_5$fitstats
fixef12 <- step12_5$fixef
cond12 <- step12_5$condition_contrast_results
time12 <- step12_5$timing_results
loc12 <- step12_5$location_contrast_results

diag12 <- step12_4$diag_summary
```

## 3. Data and model framework carried across analyses

Across scripts, model comparison was done by varying one lever at a time while attempting to keep mechanisms stable where possible.

Common model framework (where applicable):

- Outcome on log scale.
- Mixed-effects model with random intercept for `ParticipantID`.
- Additive fixed effects: `Condition + Timing_new + SampleLocation`.
- Heteroscedastic residual structure by location when estimable.
- Main inferential target: pairwise **Condition** contrasts.

## 4. Baseline non-imputed boundary analysis (Script 10)

### 4.1 Objective

Script 10 re-established a non-imputed baseline under multiple response definitions:

- PAH set sizes: top1, top4, top6, top15.
- Response modes: all-data with \(\log(x + c)\), and detects-only with \(\log(x)\).

### 4.2 Model comparability tiers

```{r}
qc10 |>
  count(comparison_tier, variance_structure, fit_success, name = "n_runs") |>
  arrange(comparison_tier)
```

Interpretation note: Script 10 included mixed comparability tiers (heteroscedastic primary fits and homoscedastic fallback fits).

### 4.3 Baseline boundary summaries

```{r}
step10_3$condition_boundary
```

```{r}
step10_3$timing_boundary
```

### 4.4 Key baseline takeaway

- Direction of the larger condition contrasts was generally stable.
- The smaller **OL vs SL** contrast showed weaker/less stable evidence.
- Timing direction was generally stable, with weaker significance in sparse settings.

## 5. Imputation beta sensitivity (Script 09)

### 5.1 Objective

Script 09 evaluated sensitivity of contrasts to imputation scaling choices while holding PAH set and model structure fixed.

### 5.2 Condition contrasts across beta settings

```{r}
cond09 |>
  select(beta_label, contrast, estimate, p.value, significant) |>
  arrange(contrast, beta_label)
```

### 5.3 Timing contrasts across beta settings

```{r}
time09 |>
  select(beta_label, contrast, estimate, p.value, significant) |>
  arrange(beta_label)
```

### 5.4 Key beta-sensitivity takeaway

- Main directional conclusions were robust to beta variation.
- Smaller contrasts showed more sensitivity in magnitude/statistical significance.

## 6. PAH-set sensitivity under consistent MLE imputation (Script 11)

### 6.1 Objective

Script 11 isolated sensitivity to the number of PAHs included in total exposure:

- top1 through top10, plus top15.
- Consistent imputation mechanism (PAH-level MLE beta).
- GM collapse and heteroscedastic mixed model retained.

### 6.2 Condition boundary summary

```{r}
step11_7$condition_boundary
```

### 6.3 Timing boundary summary

```{r}
step11_7$timing_boundary
```

### 6.4 Sensitivity figure

```{r}
step11_8$p_step8_combined
```

### 6.5 Key PAH-set takeaway

- Contrast directions remained broadly stable across PAH-set size.
- Contrast magnitudes shifted as more low-detect/imputed analytes were included.

## 7. Primary selected analysis (Script 12)

### 7.1 Final decision set

Primary analysis decisions:

1. Response uses **Top 4 PAHs**.
2. Non-detects imputed via **PAH-level MLE beta** (`elnormCensored`).
3. Repeated measures handled via **GM collapse** within timing/location/participant strata.
4. Main model is **heteroscedastic LME**.

### 7.2 Top-4 definition check

```{r}
step12_1$top4_map
```

### 7.3 Imputation summary (Top 4)

```{r}
step12_2$imputation_log |>
  select(pah_name, beta_mle, n_detect, n_cens, fit_ok, fit_method)
```

### 7.4 Collapse and model fit checks

```{r}
step12_3$step3_qc
```

```{r}
step12_4$step4_qc
```

### 7.5 Primary fit statistics and fixed effects

```{r}
fit12
```

```{r}
fixef12 |>
  select(term, estimate, std_error, df, t_value, p_value, fold_change, pct_change)
```

### 7.6 Primary contrasts

#### Condition contrasts (primary)

```{r}
cond12 |>
  select(contrast, estimate, SE, df, t.ratio, p.value, significant)
```

#### Timing contrast (secondary)

```{r}
time12 |>
  select(contrast, estimate, SE, df, t.ratio, p.value, significant)
```

#### Sampling location contrasts (exploratory)

```{r}
loc12 |>
  select(contrast, estimate, SE, df, t.ratio, p.value, significant) |>
  arrange(p.value)
```

### 7.7 Primary figures

```{r}
step12_6$p_condition_step6
```

```{r}
step12_6$p_timing_step6
```

```{r}
step12_6$p_location_step6
```

## 8. Model diagnostics (Script 12)

### 8.1 Numeric diagnostics

```{r}
diag12
```

### 8.2 Diagnostic plots

```{r}
step12_4$p_diag_combined
```

Diagnostic interpretation (brief): no major structural violations were evident; mild tail behavior and sparse location-specific outliers remain expected caveats.

## 9. Synthesis and forward manuscript path

### 9.1 What appears robust

- Major condition-direction findings are stable across sensitivity analyses.
- Timing-direction finding is generally stable.
- Primary model results are consistent with exploratory boundary analyses.

### 9.2 Where uncertainty remains

- Smaller between-condition differences (particularly OL vs SL) are less stable.
- Location-specific contrasts are exploratory due to sparse support and many pairwise tests.

### 9.3 Proposed manuscript framing

- Present Script 12 as the pre-specified/selected primary model.
- Use Scripts 09, 10, and 11 to document robustness and bounds.
- Explicitly state that inference is conditional on single imputation (imputation uncertainty not propagated).
