---
title: "Technical Analysis Update: PAH Exposure Modeling Sensitivity and Primary Model"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
---

## 1. Background and goals

This report summarizes the model-development workflow used to evaluate condition effects on total PAH exposure under sparse data and substantial non-detects.

The primary methodological decisions addressed were:

1. How to handle non-detects.
2. How many PAHs to include in the response definition.
3. How to handle repeated measures.
4. What model form to use for adjusted between-condition contrasts.

The sequence below follows the exploratory-to-primary progression:

- non-imputed baseline boundary analysis.
- sensitivity to imputation beta.
- sensitivity to number of PAHs under consistent MLE imputation.
- selected primary analysis.

There are a few adjustments made to the data:

1. The Timing variable is condensed into two levels: "Donning/Firefighting" and "Doffing"
2. As a result of this change, some participants would have more than 1 PAH measurement per timing/condition/location cell. To address this, we collapsed to the geometric mean within each cell for each participant. This changes the total number of data points from 448 to 418.
3. The single "Shirt" sample location was combined with "Sleeve".

## 2. Setup and artifact loading
- Code not shown here.

```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(kableExtra)

# Script outputs
out_09 <- "09_output"
out_10 <- "10_output"
out_11 <- "11_output"
out_12 <- "12_output"

# ---- Script 10 artifacts ----
step10_2 <- readRDS(file.path(out_10, "step2_artifacts.rds"))
step10_3 <- readRDS(file.path(out_10, "step3_artifacts.rds"))
qc10 <- readRDS(file.path(out_10, "step2_model_qc.rds"))

cond10 <- step10_2$condition_contrast_results
time10 <- step10_2$timing_results

# ---- Script 09 artifacts ----
step09_2 <- readRDS(file.path(out_09, "step2_results.rds"))
cond09 <- readRDS(file.path(out_09, "condition_contrast_results.rds"))
time09 <- readRDS(file.path(out_09, "timing_results.rds"))

# ---- Script 11 artifacts ----
step11_6 <- readRDS(file.path(out_11, "step6_artifacts.rds"))
step11_7 <- readRDS(file.path(out_11, "step7_artifacts.rds"))
step11_8 <- readRDS(file.path(out_11, "step8_artifacts.rds"))

cond11 <- step11_6$condition_contrast_results
time11 <- step11_6$timing_results

# ---- Script 12 artifacts ----
step12_1 <- readRDS(file.path(out_12, "step1_artifacts.rds"))
step12_2 <- readRDS(file.path(out_12, "step2_artifacts.rds"))
step12_3 <- readRDS(file.path(out_12, "step3_artifacts.rds"))
step12_4 <- readRDS(file.path(out_12, "step4_artifacts.rds"))
step12_5 <- readRDS(file.path(out_12, "step5_artifacts.rds"))
step12_6 <- readRDS(file.path(out_12, "step6_artifacts.rds"))

fit12 <- step12_5$fitstats
fixef12 <- step12_5$fixef
cond12 <- step12_5$condition_contrast_results
time12 <- step12_5$timing_results
loc12 <- step12_5$location_contrast_results

# Load diagnostics from dedicated files (saved in Step 7 append)
diag12 <- readRDS(file.path(out_12, "diag_summary.rds"))
p_diag_combined_12 <- readRDS(file.path(out_12, "p_diag_combined.rds"))
```

## 3. Data and model framework carried across analyses

Across scripts, model comparison was done by varying one lever at a time while attempting to keep mechanisms stable where possible.

Common model framework (where applicable):

- Outcome (Sum of PAHs, ug/m3)on log scale.
- Mixed-effects model with random intercept for `ParticipantID`.
- Additive fixed effects: `Condition + Timing_new + SampleLocation`.
- Heteroscedastic residual structure by location when estimable.
- Main inferential target: pairwise **Condition** and **Timing** contrasts.

## 4. Baseline non-imputed boundary analysis (Script 10)

### 4.1 Objective

Re-established a non-imputed baseline under multiple response definitions:

- PAH set sizes: top1 (>10% detection), top4 (>5% detection), top6 (mirroring underglove method), top15 (all PAHs).
- Response modes: all-data with $\\log(x + c)$, and detects-only with $\\log(x)$.

The data were modeled as-is with the framework described above. This provided a reference point for how the data "look" without imputation and how model fits behaved across response definitions. These results are not intended for direct inference but rather to contextualize the sensitivity analyses (beta sensitivity and PAH-set sensitivity) and the primary model. This could be considered a one side of a boundary analysis, with the other side being the fully imputed model of all 15 PAHs (worst-case, non-zero exposure estimates for ALL PAHs).

### 4.2 Model comparability tiers

```{r}
qc10 |>
  count(comparison_tier, variance_structure, fit_success, name = "n_runs") |>
  arrange(comparison_tier) |>
  kable()
```

Interpretation note: includes mixed comparability tiers (heteroscedastic primary fits and homoscedastic fallback fits). The heteroscedastic fits (preferred for these data) generally succeeded for smaller PAH sets (detects only) but struggled with larger sets (nondetects included), which is consistent with the increased sparsity and imputation challenges. 

The different variance structures aren't directly comparable, but they still provide useful context for how the data behave under different response definitions and modeling assumptions.

### 4.3 Baseline boundary summaries

```{r}
step10_3$condition_boundary |>
  kable() |>
  scroll_box(width = "100%")
```

```{r}
step10_3$timing_boundary |>
  kable() |>
  scroll_box(width = "100%")
```

### 4.4 Key baseline takeaway

- Direction of the larger condition contrasts was generally stable.
- The smaller **OL vs SL** contrast showed weaker/less stable evidence.
- Timing direction was generally stable, with weaker significance in sparse settings.

## 5. Imputation beta sensitivity (Script 09)

### 5.1 Objective

Evaluated sensitivity of contrasts to imputation scaling choices while holding PAH set and model structure fixed.

### 5.2 Condition contrasts across beta settings

```{r}
cond09 |>
  select(beta_label, contrast, estimate, p.value, significant) |>
  arrange(contrast, beta_label) |>
  kable()
```

- Interpretation: the main directional conclusions (SS > SL > OL) were consistent across beta settings, but the statistical significance of contrasts varied more for smaller contrasts (OL vs SL) than for larger contrasts (SS vs SL, SS vs OL). This suggests that while the overall pattern of differences is robust to imputation scaling, the confidence in smaller contrasts is more sensitive to how non-detects are handled.

### 5.3 Timing contrasts across beta settings

```{r}
time09 |>
  select(beta_label, contrast, estimate, p.value, significant) |>
  arrange(beta_label) |>
  kable()
```

- Interpretation: the timing contrast (Doffing > Donning/Firefighting) was generally consistent in direction and statistical significance across beta settings. The magnitude of the contrast varied with the choice of beta, with small betas having larger estimated contrasts and stronger significance, while larger betas had smaller estimated contrasts and weaker significance. This suggests that the timing effect is somewhat sensitive to imputation scaling, but the overall direction of the effect is robust.

### 5.4 Key beta-sensitivity takeaway

- Main directional conclusions were robust to beta variation.
- Smaller contrasts showed more sensitivity in magnitude/statistical significance.

## 6. PAH-set sensitivity under consistent MLE imputation (Script 11)

### 6.1 Objective

Evaluated sensitivity to the number of PAHs included in total exposure:

- top1 through top10, plus top15.
- Consistent imputation mechanism (PAH-level MLE beta).
- GM collapse and heteroscedastic mixed model retained.

### 6.2 Condition boundary summary

```{r}
step11_7$condition_boundary |>
  kable()
```

- For all evaluated PAH sets, the largest contrast (SS vs OL) remained significant and in the same direction, with SS associated with higher average exposure. The SS vs SL contrast also remained significant and in the same direction, again with SS associated with a higher average exposure. The smaller contrast (OL vs SL) showed more variability in significance across PAH sets, with some sets showing significant differences and others not - this was associated with the number of PAHs (top 8, 9, 10, 15 were significant while top 7 or fewer were not). This suggests that the main condition effect (SS > OL) is robust to the number of PAHs included, while the smaller contrast (OL vs SL) is more sensitive to this choice.

### 6.3 Timing boundary summary

```{r}
step11_7$timing_boundary |>
  kable()
```

- For all evaluated PAH sets, the timing contrast (Doffing > Donning/Firefighting) remained significant and in the same direction, with Doffing associated with higher average exposure. The magnitude of the contrast varied across PAH sets, with some sets showing larger estimated contrasts than others, but the overall pattern of significance and direction was consistent. Roughly, as more PAHs were included, the estimated difference decreased. This suggests that the timing effect is robust to the number of PAHs included in the total exposure definition.

### 6.4 Sensitivity figure

```{r}
step11_8$p_step8_combined
```

### 6.5 Key PAH-set takeaway

- Contrast directions remained broadly stable across PAH-set size.
- Contrast magnitudes shifted as more low-detect/imputed analytes were included.

## 7. Primary selected analysis (Script 12)

### 7.1 Final decision set

Primary analysis decisions:

1. Response uses **Top 4 PAHs**.
2. Non-detects imputed via **PAH-level MLE beta** (`elnormCensored`).
3. Repeated measures handled via **GM collapse** within timing/location/participant strata.
4. Main model is **heteroscedastic LME**.

### 7.2 Top-4 definition check

```{r}
step12_1$top4_map |>
  kable()
```

### 7.3 Imputation summary (Top 4)

```{r}
step12_2$imputation_log |>
  select(pah_name, beta_mle, n_detect, n_cens, fit_ok, fit_method) |>
  kable()
```

### 7.4 Collapse and model fit checks

```{r}
as_tibble(step12_3$step3_qc, .name_repair = "unique") |>
  kable()
```

```{r}
step12_4$step4_qc |>
  kable()
```

### 7.5 Primary fit statistics and fixed effects

```{r}
fit12 |>
  kable()
```

```{r}
fixef12 |>
  select(term, estimate, std_error, df, t_value, p_value, fold_change, pct_change) |>
  kable()
```

### 7.6 Primary contrasts

#### Condition contrasts (primary)

```{r}
cond12 |>
  select(contrast, estimate, SE, df, t.ratio, p.value, significant) |>
  kable()
```

- Interpretation: the main condition contrasts (SS vs SL and SS vs OL) were significant and in the expected direction, with SS associated with higher average exposure (SS > SL > OL). The smaller contrast (OL vs SL) was not significant, suggesting less confidence in a difference between these two conditions. This pattern is consistent with the boundary analyses and sensitivity analyses, reinforcing the robustness of the main condition effect while highlighting uncertainty in the smaller contrast.

#### Timing contrast (secondary)

```{r}
time12 |>
  select(contrast, estimate, SE, df, t.ratio, p.value, significant) |>
  kable()
```

- Interpretation: the timing contrast (Doffing > Donning/Firefighting) was significant and in the expected direction, with Doffing associated with higher average exposure. This is consistent with the boundary analyses and sensitivity analyses, reinforcing the robustness of the timing effect.

#### Sampling location contrasts (exploratory)

```{r}
loc12 |>
  select(contrast, estimate, SE, df, t.ratio, p.value, significant) |>
  arrange(p.value) |>
  kable()
```

### 7.7 Primary figures

```{r}
step12_6$p_condition_step6
```

```{r}
step12_6$p_timing_step6
```

```{r}
step12_6$p_location_step6
```

## 8. Model diagnostics (Script 12)

### 8.1 Numeric diagnostics

```{r}
diag12 |>
  kable()
```

### 8.2 Diagnostic plots

```{r}
p_diag_combined_12
```

Diagnostic interpretation (brief): no major structural violations were evident; mild tail behavior and sparse location-specific outliers remain expected caveats.

## 9. Synthesis and forward manuscript path

### 9.1 What appears robust

- Major condition-direction findings are stable across sensitivity analyses (SS associated with higher average exposure).
- Timing-direction finding is generally stable (Doffing associated with higher average exposure).
- Primary model results are consistent with exploratory boundary analyses.

### 9.2 Where uncertainty remains

- Smaller between-condition differences (particularly OL vs SL) are less stable.
- Location-specific contrasts are exploratory due to sparse support and many pairwise tests.

### 9.3 Proposed manuscript framing

- Present selected primary model.
- Document robustness and bounds.
- Explicitly state that inference is conditional on single imputation (imputation uncertainty not propagated).
